---
title: "Reproducible Research: Peer Assessment 1"
author: "HT"
date: "May 21, 2016 and Jul 17,2016"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data
```{r}
activity_data <- read.csv("activity.csv", header = TRUE, na.strings = "NA")
summary(activity_data)
```

## What is mean total number of steps taken per day?
```{r}
total_steps_per_day <- tapply(activity_data$steps, activity_data$date, sum, na.rm = TRUE)
hist(total_steps_per_day, main = "Frequency of number of steps per day", 
     xlab = "Number of steps per day", ylab = "Frequency", col = "blue")
mean_total_number_of_steps_per_day = tapply(activity_data$steps, activity_data$date, mean, na.rm=TRUE)
mean_total_number_of_steps <- mean(total_steps_per_day, na.rm = TRUE)
median_total_number_of_steps_per_day <- median(total_steps_per_day, na.rm = TRUE)

print(paste("Mean total number of steps = ",mean_total_number_of_steps))
print(paste("Median total number of steps per day = ", median_total_number_of_steps_per_day ))
```


## What is the average daily activity pattern?   
``` {r}
mean_steps_per_interval <- tapply(activity_data$steps, activity_data$interval, mean, na.rm = TRUE)
plot(mean_steps_per_interval, type="l",main="Average Daily Activity Pattern", 
     xlab="Daily Interval", ylab="Mean Total Number of Steps per day")
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
``` {r}
max_number_steps_sequence <- seq(along = mean_steps_per_interval)[mean_steps_per_interval == max(mean_steps_per_interval)]
max_number_steps_sequence

# Get the steps data as a vector
tmp_dataTest <- as.vector(activity_data$steps)
# Set it to one where data is missing
tmp_dataTest[is.na(tmp_dataTest)] = 1

```
### Total number of rows with missing values   

``` {r}
##sum(is.na(activity_data$steps))
print(paste("Total number of rows in data set = ", nrow(activity_data)))
print(paste("Total number of rows with complete data (i.e. no missing data in any field = )", sum(complete.cases(activity_data))))
print(paste("Total number of rows with missing values = ",nrow(activity_data) - sum(complete.cases(activity_data))))
```

## Imputing missing values   
```{r}
## We test each element of 'steps'; if it is NA, we replace with the mean, otherwise we replace with the original value.
imputed_data = transform(activity_data, steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))

stepsTotalPerDay_NoMissing <- tapply(imputed_data$steps, imputed_data$date, sum)
hist(stepsTotalPerDay_NoMissing, breaks = 6, main = "Frequency of number of steps per day", xlab = "Number of steps per day", ylab = "Frequency", col = "blue")
## Mean of imputed data
imputedmean_total_number_of_steps_per_day = tapply(imputed_data$steps, imputed_data$date, mean, na.rm=TRUE)
imputedmean_total_number_of_steps <- mean(stepsTotalPerDay_NoMissing, na.rm = TRUE)
imputedmedian_total_number_of_steps_per_day <- median(stepsTotalPerDay_NoMissing, na.rm = TRUE)

print(paste("Mean total number of steps = ",imputedmean_total_number_of_steps))
print(paste("Median total number of steps per day = ", imputedmedian_total_number_of_steps_per_day ))

if(imputedmean_total_number_of_steps == imputedmedian_total_number_of_steps_per_day) {
  print("The impact of imputing the missing steps data is that, the Mean and median of imputed data are equal, suggesting a standard normal distribution. This is confirmed by the histogram below.")
}
hist(stepsTotalPerDay_NoMissing, main = "Frequency of number of steps per day for imputed data", 
     xlab = "Number of steps per day", ylab = "Frequency", col = "blue")

```


## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Create a factor variable with two levels (weekday, weekend-day)
LT <- as.POSIXlt(imputed_data$date, format = "%Y-%m-%d")
WeekDays <- LT$wday

WeekDays[WeekDays == 0] = 0 ## Sunday
WeekDays[WeekDays == 6] = 0 ## Saturday
WeekDays[WeekDays != 0] = 1 ## Other week day

WeekDaysFactor <- factor(WeekDays, levels = c(0, 1))
# Add the factor variable to the data
imputed_data$WD <- WeekDaysFactor
# Calculate the mean
MeanStepsPerWeekday <- tapply(imputed_data$steps, list(imputed_data$interval, imputed_data$WD), mean, 
    na.rm = TRUE)

par(mfrow = c(2, 1))
# Display the 2 plots
with(imputed_data, {
    par(mai = c(0, 1, 1, 0))
    plot(MeanStepsPerWeekday[, 1], type = "l", main = ("Steps vs. Interval"), 
        xaxt = "n", ylab = "Week-ends")
    title = ("# of Steps v.s. Interval")
    par(mai = c(1, 1, 0, 0))
    plot(MeanStepsPerWeekday[, 2], type = "l", xlab = "Interval", ylab = "Week days")

})
```
## Showing difference in the plots in terms of stats too (apart from the visual above)
```{r}
## These are the summary stats of week-ends
summary(MeanStepsPerWeekday[, 1])
## This is the Standard Deviation of the week-end data
sd(MeanStepsPerWeekday[, 1])
## These are the summary stats of weekdays
summary(MeanStepsPerWeekday[, 2])
## This is the Standard Deviation of the weekday data
sd(MeanStepsPerWeekday[, 2])
## Result: Therefore, based on the above visual plot and summary statistics presented above:
## we infer that there is some difference between activity on weekdays and week-ends
## but not a whole lot (i.e. not highly significant difference)
## However, it is noted that, the Mean number of steps in the week-end is higher than the Mean number of steps in the week days.
## This suggests, more people walk (more) in the week-ends than weekdays.
## Also, the maximum number of steps walked in the week days is higher than the maximum number of steps walked in the week-ends.
## This could possibly be because of some out-liers in data wherein, someone with an outdoors job (e.g. delivery) may be causing this increase, whereas, people may be going out on shorter walks in the week-ends, which seems to reduce the maximum number of steps walked in the week-ends.
```